Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2025-06-25 paper/main.tex
include GPfit description and post mean details

To handle the functional nature of this output, we reformulate the problem using principal
components (PCs; \textcolor{red}{@Steve - can you add a textbook reference?}).  
We first center all $\hat{S}_j$ by subtracting $\bar{S} = \frac{1}{111}\sum_{j=1}^{111} \hat{S}_j$, then
we combine them into an $n\times m$ matrix $\boldsymbol\eta$.  We use principal components to decompose 
$\boldsymbol\eta = BW^\top$ where $B$ is the $n\times p_\eta$ matrix of basis functions and $W$
is the $m\times p_\eta$ matrix of basis weights.  Details on this decomposition are 
reserved for \ref{sec:apdx_basis}.  For the CAMB dataset, $p_\eta=6$; for Mira-Titan, $p_\eta=8$.
As a visual, the left panel of Figure \ref{fig:mean_PCs_oneW} shows $\bar{S}$ for the Mira-Titan 
training cosmologies.  The center panel shows the $p_\eta=8$ basis functions.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{mean_PCs_oneW.png}
    \caption{{\it Left:} The estimated mean trend of the $m=111$ posterior means from 
    the Mira-Titan dataset. {\it Middle:} The $p_\eta$ principal components obtained from 
    the posterior means. {\it Right:} An 
    illustration of how the weights for the first PC will change as the first 
    cosmological parameter varies \textcolor{red}{This last plot is wonky because 
    it was made with only the $N=6$ test cosmologies, not a large uniform or FF design, 
    Steve can update this}.
    \textcolor{red}{This figure also needs work.  Axes need to be fixed.  Caption needs re-write.
    Right panel needs to be redone.  It should look smooth like a GP.  Or else just remove it.}}
    \label{fig:mean_PCs_oneW}
\end{figure}

Let $w_i^{(j)}$ denote the $ij^\textrm{th}$ element of $W$, representing the basis weight for the 
$j^\textrm{th}$ cosmology and the $i^\textrm{th}$ PC.
Conditional on $B$, our training data now consists of scalars $w_i^{(j)}$ as a function of $\psi_j$
for $j \in \{1,\ldots,m\}$ and $i \in \{1,\ldots,p_\eta\}$.  If we can effectively predict these basis weights
as a function of $\psi$ (i.e., $\hat{w}_i(\psi)$ for $i \in \{1,\ldots,p_\eta\}$), then we may obtain predictions 
of the power spectrum as
\[
\hat{S}(\psi) = \sum_{i=1}^{p_\eta} \hat{w}_i(\psi) B_i,
\]
where $B_i$ is the $i^\textrm{th}$ basis function (column) of $B$.

% This is old: switched to SEPIA
% We assume a powered exponential kernel for the Gaussian process and employ 
% \texttt{GPfit} to perform a multi-start gradient based search for GP's hyperparameters 
% \citep{macdonald2015gpfit}. 
In alignment with \cite{heitmann2009coyote} and \cite{higdon2010estcosmo}, we use 
GPs to model these principal components' weights. For direct comparison alongside
\cite{moran2023mira}, we employ a Python implementation of this approach called 
SEPIA \citep{gattiker2020sepia}. That is, each $w_i(\psi)$ is modeled as a 
zero-mean Gaussian process with a squared exponential (i.e., Gaussian) covariance function:
\[
w_i(\psi) \sim GP\left(0,\ \frac{1}{\lambda_{z_i}} R^{(i)} + \frac{1}{\lambda_{s_i}} I \right),
\]
with covariance function
\[
R^{(i)}_{uv} = \exp\left(-\sum_{k=1}^{p_\psi} \beta_{ik} (x_{uk} - x_{vk})^2 \right),
\]
where $\beta_{ik}$ is the inverse squared lengthscale for the $k$th input dimension, 
$\lambda_{z_i}$ is the marginal precision for $w_i(\psi)$, and $\lambda_{s_i}$ is the nugget 
precision (set to zero if no nugget is used).
\textcolor{red}{(@Steve - there is a lot missing here.  The inputs to the covariance kernel
here need to be $\psi$.  Are we using a nugget?  If not, just remove it.  How are these parameters
being estimated?  Make sure this kernel notation matches what we used earlier as neatly as possible.)}

Using these trained GPs from the PCs, we predict the spectra for each of the $N$ held-out 
cosmologies for both the CAMB and Mira-Titan datasets. \textcolor{red}{(@Steve - this needs work too.
Please include the formula for the GP posterior mean, which is how we obtain $\hat{w}_i$.  We never
had use to include it earlier, but we should use it here.  If you need, you can introduce new matrix/vector
notation to represent the training data in this scenario.  At this point, you should refer back to the right
panel of Figure \ref{fig:mean_PCs_oneW} and narrate how this shows the predicted GP mean for $w_1$ (for Mira-Titan)
as a function of the first cosmological parameter $\psi_1$.)}

@@

\textcolor{red}{(@Steve - revise the text so it works for any $m$, not just 111.  
Also, replace $\Gamma$ with $W$ so it matches the text I revised earlier.)}

Here, we begin with a $n \times m$ matrix composed of $m=111$ posterior means, 
each of length $n$, denoted by $\boldsymbol\eta$. To facilitate more efficient estimation, 
we use singular value decomposition \citep[SVD; e.g.,][]{banerjee2014linear}, 
where $\boldsymbol\eta = UDV^T$; $U$ and $V$ are orthogonal matrices of size $n \times n$ 
and $m\times m$, respectively. $D$ is a diagonal matrix containing the singular values. 
For this application, $\boldsymbol\eta$ has rank $r=\text{min}\{m,n\}$ (111 for Mira-Titan, 
32 for CAMB), which determines the number of non-zero elements in $D$. We find an alternative 
decomposition where $\Sigma$ is a diagonal matrix with the non-zero singular values of 
$\boldsymbol\eta$, and $U_1$ and $V_1$ are matrices with orthonormal columns of size 
$n_\eta \times r$ and $m \times r$, respectively.

>>>>

To handle the functional nature of this output, we reformulate the problem using principal
components \citep[PCs, e.g.][]{banerjee2014linear}.  
We first center all $\hat{S}_j$ by subtracting $\bar{S} = \frac{1}{m}\sum_{j=1}^{m} \hat{S}_j$, 
then we combine them into an $n\times m$ matrix $\boldsymbol\eta$.  We use the first $p_\eta$ 
principal components to decompose $\boldsymbol\eta = B\Gamma^\top$ where $B$ is the 
$n\times p_\eta$ matrix of basis functions and $\Gamma$ is the $m\times p_\eta$ matrix of basis weights.  
Details on this decomposition are reserved for \ref{sec:apdx_basis}. For both the CAMB and Mira-Titan 
datasets, we set $p_\eta=10$. As a visual, the left panel of Figure \ref{fig:mean_PCs_oneW} shows 
$\bar{S}$ for the Mira-Titan training cosmologies.  The center panel shows the $p_\eta$ basis functions.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{mean_PCs_oneW.jpeg}
    \caption{{\it Left:} The estimated mean trend of the $m=111$ posterior means from 
    the Mira-Titan dataset. {\it Middle:} The $p_\eta$ principal components obtained from 
    the posterior means. {\it Right:} An 
    illustration of how the weights for the fifth PC ($\gamma_5$) will change as the first 
    cosmological parameter ($\psi_1$) varies.}
    \label{fig:mean_PCs_oneW}
\end{figure}

Let $\gamma_i^{(j)}$ denote the $ij^\textrm{th}$ element of $\Gamma$, representing the basis weight for the 
$j^\textrm{th}$ cosmology and the $i^\textrm{th}$ PC.
Conditional on $B$, our training data now consists of scalars $\gamma_i^{(j)}$ as a function of $\psi_j$
for $j \in \{1,\ldots,m\}$ and $i \in \{1,\ldots,p_\eta\}$.  If we can effectively predict these basis weights
as a function of $\psi$ (i.e., $\hat{\gamma}_i(\psi)$ for $i \in \{1,\ldots,p_\eta\}$), then we may obtain predictions 
of the power spectrum as
\[
\hat{S}(\psi) = \sum_{i=1}^{p_\eta} \hat{\gamma}_i(\psi) B_i,
\]
where $B_i$ is the $i^\textrm{th}$ basis function (column) of $B$.

% SEPIA information:
% In alignment with \cite{heitmann2009coyote} and \cite{higdon2010estcosmo}, we use 
% GPs to model these principal components' weights. For direct comparison alongside
% \cite{moran2023mira}, we employ a Python implementation of this approach called 
% SEPIA \citep{gattiker2020sepia}. That is, each $\gamma_i(\psi)$ is modeled as a 
% zero-mean Gaussian process with a squared exponential (i.e., Gaussian) covariance function:
% \[
% \gamma_i(\psi) \sim GP\left(0,\ \frac{1}{\lambda_{z_i}} R^{(i)} + \frac{1}{\lambda_{s_i}} I \right),
% \]
% with covariance function
% \[
% R^{(i)}_{uv} = \exp\left(-\sum_{k=1}^{p_\psi} \beta_{ik} (x_{uk} - x_{vk})^2 \right),
% \]
% where $\beta_{ik}$ is the inverse squared lengthscale for the $k$th input dimension, 
% $\lambda_{z_i}$ is the marginal precision for $\gamma_i(\psi)$, and $\lambda_{s_i}$ is the nugget 
% precision (set to zero if no nugget is used).
% \textcolor{red}{(@Steve - The inputs to the covariance kernel
% here need to be $\psi$.  Are we using a nugget?  If not, just remove it.  How are these parameters
% being estimated?  Make sure this kernel notation matches what we used earlier as neatly as possible.)}

In alignment with \cite{higdon2010estcosmo} and \cite{moran2023mira}, we use 
GPs to model the principal components' weights. 
We assume a powered exponential kernel for the Gaussian process and employ 
\texttt{GPfit} to perform a multi-start gradient based search for GP's hyperparameters 
\citep{macdonald2015gpfit}. That is, the $i$th vector of PC weights $\gamma_i(\psi)$, 
$i \in \{1,\ldots,p_\eta\}$, will be modeled as a zero mean GP: 
$\gamma_i(\psi) \sim \mathrm{GP}(0, \sigma^2R)$, with

\begin{equation*}
    R_{uv} = \prod_{j=1}^{p_\psi}\exp\left\{-10^{\beta_j}|\psi_{uj}-\psi_{vj}|^\alpha\right\}.
\end{equation*}

We fix $\alpha=1.95$ and estimate $\sigma^2$ and each $\beta_j$ with $j \in \{1,...,p_\psi\}$, and repeat this
process for all of the $p_\eta$ PCs. Using these trained GPs from the PCs, we predict the spectra for 
each of the $N$ held-out cosmologies (such as $\psi^*$) for both the CAMB and Mira-Titan datasets.

We fix the power parameter $\alpha = 1.95$ and estimate the lengthscale parameters $\beta_1, \ldots, \beta_{p_\psi}$ 
from the data using maximum likelihood. To obtain the predicted basis weight $\hat{\gamma}_i(\psi^*)$ 
for a new cosmology $\psi^* \in \mathbb{R}^{p_\psi}$, 
we use the posterior mean from a zero-mean GP model fit to the training data 
$\{(\psi_j, \gamma_i^{(j)})\}_{j=1}^m$. Under a zero-mean GP, the posterior mean is given by:

\begin{equation*}
\hat{\gamma}_i(\psi^*) = r(\psi^*)^\top R^{-1} \boldsymbol{\gamma}_i,
\end{equation*}

where $\boldsymbol{\gamma}_i = [\gamma_i^{(1)}, \gamma_i^{(2)}, \ldots, \gamma_i^{(m)}]^\top$ is the vector of 
$m$ observed PC weights for the $i$th component, $R$ is the $m \times m$ correlation matrix between 
all pairs of training inputs defined above, and $r(\psi^*) \in \mathbb{R}^m$ is the vector of correlations between 
the test input $\psi^*$ and each training input $\psi_j$. To illustrate, in the right panel of Figure \ref{fig:mean_PCs_oneW}, 
we show how the predicted GP mean for the weight of the fifth PC ($\gamma_5$) for Mira-Titan 
varies as a function of the first cosmological parameter $\psi_1$, while the other parameters are fixed at a (scaled)
midpoint value of 0.5.

@@

Here, we begin with a $n \times m$ matrix composed of $m$ posterior means, 
each of length $n$, denoted by $\boldsymbol\eta$. To facilitate more efficient estimation, 
we use singular value decomposition \citep[SVD; e.g.,][]{banerjee2014linear}, 
where $\boldsymbol\eta = UDV^T$; $U$ and $V$ are orthogonal matrices of size $n \times n$ 
and $m\times m$, respectively. $D$ is a diagonal matrix containing the singular values. 
Here, $\boldsymbol\eta$ has rank $r=\text{min}\{m,n\}$ (for our applications, $r=111$ for Mira-Titan 
and $r=32$ for CAMB), which determines the number of non-zero elements in $D$. We find an alternative 
decomposition where $\Sigma$ is a diagonal matrix with the non-zero singular values of 
$\boldsymbol\eta$, and $U_1$ and $V_1$ are matrices with orthonormal columns of size 
$n_\eta \times r$ and $m \times r$, respectively.
